Combinatorics of the geometry of Wilson loop diagrams II: Grassmann necklaces, dimensions, and denominators

### 2019-10-16 GWLDCII.tex
rewrote lemma 4.2, plus minor edits throughout

As noted above, the sum of the Feynman integrals associated to Wilson loop diagrams gives the volume of a convex polytope. if the singularities appearing in individual integrals cancel out in the overall sum, then the overall volume is finite. This cancellation has been conjectured in general \cite{Hodges2013} and verified explicitly in a few cases \cite{casestudy, HeslopStewart}, but it is hard to prove in general due to the fact that the relationship between Wilson loop diagrams and $3k$-dimensional positroid cells of $\Gr(k,n)$ is neither one-to-one nor onto. In this paper, we give a characterization of the factors appearing in the denominator of these Feynman integrals in terms of the Grassmann necklace of the associated positroid, thus providing a much simpler and more tractable link between the singularities and the diagrams.

@@

In this section, we give an algorithm for passing directly from the Wilson loop diagrams to its Grassmann necklace. This not only greatly simplifies the previously known process to obtain a positroid cell from a Wilson loop diagram, but will also allow us to relate the behavior of the positroid $M(W)$ directly to the configuration of propagators in $W$. In section \ref{sec poles} we see that the denominator, $R(W)$, of the integrand associated to a Wilson loop diagram, $\Omega(W)$, can be written in terms of the Grassmann necklace of a diagram.

@@

\begin{lem}\label{lem dihedral}
If two Wilson loop diagrams differ by a dihedral transformation then their Le diagrams have same number of plusses.
\end{lem}
\begin{proof}
By \cite[Proposition 17.10]{Postnikov}, the dimension of a positroid (and hence the number of plusses in its Le diagram) is $k(n-k) - A(\pi_W)$, where $A(\pi_W)$ denotes the number of alignments of the decorated permutation $\pi_W$ of the positroid associated to $W$. (See \cite[Figure 17.1]{Postnikov} and preceding discussion.)

It can easily be seen from \cite[Section 17]{Postnikov} and Algorithm \ref{alg:put GN on WLD} that dihedral transformations of a Wilson loop diagram $W$ correspond to dihedral transformations of the chord diagram representation of $\pi_W$. Since the number of alignments in a chord diagram is preserved under dihedral transformations, the result follows.
\end{proof}

@@

In this section we investigate the poles of the integrand for $I(W)$. Since SYM $N=4$ is a finite theory \cite{Alday:2007hr, ParkeTaylorids}, the scattering amplitudes (and thus the holomorphic Wilson loops $\,\mathfrak{W}_{k,n}$) are finite. That is, for a fixed set of external particles $\cZ$, we must have
\bas \sum_{W \text{ s.t. } |\cP| = k} I(W)(\cZ) <\infty \;.\eas
This holds if all poles of the integrands cancel. As a step towards this, in this section we prove some results about the structure of the denominator $R(W)$ of the integrand, as defined in Definition~\ref{def R(W)}.

The results of Section~\ref{sec GN algorithm} allow us to relate the configuration of propagators in a Wilson loop diagram $W$ to minors of $C(W)$ that contribute to $R(W)$.

In this section, we fix an arbitrary ordering on the propagators of W, i.e we impose an ordering on the rows of $C(W)$.

The main result of this section is Theorem~\ref{thm denom}, which expresses the denominator $R(W)$ in terms of the Grassmann necklace of $W$. This simplifies the computation of $R(W)$ and allows us to directly relate the poles of the integrand to the combinatorics of the diagram.

@@

That is, each $r_i$ is the determinant of the minor determined by the propagators (rows) that contribute different values (columns) to the $i^{th}$ element of the Grassmann necklace than they did to the $i-1^{st}$. Since each $r_i$ is simply the determinant of a particular submatrix of $C(W)$ and an ordering on the propagators has been fixed, the sign of each $r_i$ is well-defined.

@@

Below, we show that the polynomial $R$ associated to Wilson loop diagram $W$ is equal to the denominator $R(W)$ from Definition~\ref{def R(W)}, and is the radical of the product $\prod_{i=1}^n \Delta_{I_i}$. First, however, we give a worked example of Algorithm~\ref{alg WLD to denom via GN}.

@@

\begin{eg}
Consider the Wilson loop diagram in Figure~\ref{fig R eg}. Assigning propagators $p$, $q$, $r$ to rows $1,2,3$ respectively, we obtain the matrix
\[
C(W) = \begin{bmatrix} a & b & 0 & 0 & 0 & c & d \\ e & f & 0 & 0 & g & h & 0 \\ i & j & 0 & k & l & 0 & 0 \end{bmatrix}
\]
The Grassmann necklace of this diagram is 
\begin{gather*}I_1 = \{1,2,4\}, I_2 = \{2,4,5\}, I_3 = \{4,5,6\}, I_4=\{4,5,6\},\\ I_5=\{5,6,7\}, I_6 = \{6,7,1\}, I_7=\{7,1,2\}. \end{gather*}  
Figure~\ref{fig R eg} indicates indicates which vertex is contributed to $I_i$ by each propagator, for each $i \in [1,7]$.

@@

All that remains is to prove that $\Delta_{I_i}$ has no irreducible factors of degree 3 or more.  Suppose for a contradiction that $f$ is a factor of $\Delta_{I_i}$ of degree $\geq 3$. Without loss of generality\footnote{This is without loss of generality for the following reason.  By removing the propagators which come before those contributing to $f$ (in the order given by $I_i$) and changing $i$ to be the first vertex which contributes to $f$, we obtain an admissible diagram.  In this diagram, $f$ still divides $\Delta_{I_i}$ and we also have $i\in I_i$ and that 
$i$ contributes to $f$.  Thus this diagram can be used in place of $W$.} we may assume $i\in I_i$ and that $i$ contributes to $f$.
Finally, we can suppose that $W$ is minimal in number of propagators with an irreducible factor of degree $\geq 3$.

@@

\textbf{Case 2}: Suppose $p = (i, m)$, i.e. $p$ has an end on the $i^{th}$ edge. If no other propagators are supported on the vertex $i$ then the corresponding column of $C(W)$ has only one nonzero entry in it, and so $\Delta_{I_i}$ has a linear factor contributed by $p$ and $i$; as above, this is a contradiction.
Since $p$ is the first propagator in the ordering imposed by $I_i$, it is the clockwise-most propagator supported on $i$ (by Remark {\ref{clockwise ordering rem}}). Since there are multiple propagators supported on $i$, there must be a $q$ ending on edge $i$, such that $I_i(q) = i+1$, that is second clockwise-most propagators supported by $i$ (see Figure~\ref{fig no big factors}).
Write $q = (i, j)$ for the support of $q$.  The situation for $q$ is very similar to case 1: in particular, we have $I_{i+1}(q) = j$ by Lemma {\ref{vertex cyclic int lem}} and so if $j\in I_i$ then the propagator which contributes $j$ is outside of $q$.
    %\color{violet}if we have $r$ such that $I_i(r) = j$, as above, $r$ lies outside $q$: $r \in \cP_{out}(q, <_i)$.

@@

\item Suppose propagators $p$ and $q$ are supported on a common edge  $a$, with $I_i(p)=a$ and $I_i(q)=a+1$.  Let $x_{p,a},x_{p,a+1},x_{q,a},x_{q,a+1}$ be the associated variables in $C(W)$. For any fixed bijection $\sigma$ from $\cP\setminus\{p,q\}$ to $I_i \setminus\{a,a+1\}$ for which each propagator is supported on its image under the bijection, we can extend $\sigma$ to a bijection from propagators to $I_i$ in two ways: either $p\mapsto a$ and $q\mapsto a+1$ or $p\mapsto a+1$ and $q\mapsto a$.  The sum of the contributions of all these bijections to $\Delta_{I_i}$ is therefore the product of $(x_{p,a}x_{q,a+1}-x_{p,a+1}x_{q,a})$  with the minor coming from $\cP\setminus\{p, q\}$ and $I_i \setminus \{a,a+1\}$.  Since there is no cancellation of terms in the expansion of $\Delta_{I_i}$, if any other terms appear then they will cause a factor which is not in the form described at the beginning of the proof of part (2).  Therefore no such terms exist and $(x_{p,a}x_{q,a+1}-x_{p,a+1}x_{q,a})$ is a factor of $\Delta_{I_i}$.

@@

Taking everything together we have that $R|\prod_{i=1}^n \Delta_{I_i}$, $R$ contains all factors of $\prod_{i=1}^n \Delta_{I_i}$ and $R$ is square-free.  Therefore the ideal generated by $R$ is the radical of the ideal generated by $\prod_{i=1}^n \Delta_{I_i}$.
  \end{enumerate}
\end{proof}

@@

Thus we have shown a deep relationship between the physically derived poles of the integrand $\Omega(W)$ and the geometry and combinatorics associated to the diagrams $W$. This is yet another piece of substantial evidence indicating that the positroid representation of Wilson loop diagrams is the correct geometric object to be considering when considering the Wilson loop amplitudes geometrically.

>>>>

As noted above, the sum of the Feynman integrals associated to Wilson loop diagrams gives the volume of a convex polytope. If the singularities appearing in individual integrals cancel out in the overall sum, then the overall volume is finite. This cancellation has been conjectured in general \cite{Hodges2013} and verified explicitly in a few cases \cite{casestudy, HeslopStewart}, but it is hard to prove in general due to the fact that the relationship between Wilson loop diagrams and $3k$-dimensional positroid cells of $\Gr(k,n)$ is neither one-to-one nor onto. In this paper, we give a characterization of the factors appearing in the denominator of these Feynman integrals in terms of the Grassmann necklace of the associated positroid, thus providing a much simpler and more tractable link between the singularities and the diagrams.

@@

In this section, we give an algorithm for passing directly from the Wilson loop diagrams to its Grassmann necklace. This not only greatly simplifies the previously known process to obtain a positroid cell from a Wilson loop diagram, but will also allow us to relate the behavior of the positroid $M(W)$ directly to the configuration of propagators in $W$. In section \ref{sec poles} we use these results to show that the denominator $R(W)$ of the integrand associated to a Wilson loop diagram can be written in terms of the Grassmann necklace of a diagram.

@@

\begin{lem}\label{lem dihedral}
If two Wilson loop diagrams differ by a dihedral transformation then their Le diagrams have same number of plusses.
\end{lem}
\begin{proof}
Recall that the number of plusses in a Le diagram is given by the dimension of the corresponding positroid. By \cite[Proposition 17.10]{Postnikov}, the dimension of a positroid $M$ is $k(n-k) - A(\pi_M)$, where $A(\pi_M)$ denotes the number of alignments of the decorated permutation $\pi_M$ of $M$. The alignments of a decorated permutation are defined in terms of its chord diagram representation, where an alignment is any pair of non-crossing chords that share the same orientation. See \cite[Figure 17.1]{Postnikov} and preceding discussion for definitions and more details.

With the help of Algorithm \ref{alg:put GN on WLD}, it is now easy to trace the effect of a dihedral transformation on a Wilson loop diagram through its Grassmann necklace to its decorated permutation to its chord diagram, and see that dihedral transformations of a Wilson loop diagram $W$ correspond to dihedral transformations of the chord diagram representation of $\pi_{M(W)}$. Since the number of alignments in a chord diagram is clearly preserved under dihedral transformations, the result follows.
\end{proof}

@@

Recall from Section~\ref{section WLD background} that each Wilson loop diagram defines a functional $I(W)$, which associates a volume to the positroid cell $\Sigma(W)$. In this section we investigate the poles of the integrand of $I(W)$.

Since SYM $N=4$ is a finite theory \cite{Alday:2007hr, ParkeTaylorids}, the scattering amplitudes (and thus the holomorphic Wilson loops $\,\mathfrak{W}_{k,n}$) are finite. That is, for a fixed set of external particles $\cZ$ we must have
\bas \sum_{W \text{ s.t. } |\cP| = k} I(W)(\cZ) <\infty \;.\eas
This holds if all poles of the integrands cancel. As a step towards this, in this section we prove some results about the structure of the denominator $R(W)$ of the integrand, as defined in Definition~\ref{def R(W)}.

The results of Section~\ref{sec GN algorithm} allow us to relate the configuration of propagators in a Wilson loop diagram $W$ to minors of $C(W)$ that contribute to $R(W)$. The main result of this section is Theorem~\ref{thm denom}, which expresses the denominator $R(W)$ in terms of the Grassmann necklace of $W$. This simplifies the computation of $R(W)$ and allows us to directly relate the poles of the integrand to the combinatorics of the diagram.

For the duration of this section, fix an arbitrary ordering on the propagators of $W$. This imposes an ordering on the rows of $C(W)$.

@@

That is, each $r_i$ is the determinant of the minor determined by the propagators (rows) that contribute different values (columns) to the $i^{th}$ element of the Grassmann necklace than they did to the $(i-1)^{th}$. Since each $r_i$ is simply the determinant of a particular submatrix of $C(W)$ and an ordering on the propagators has been fixed, the sign of each $r_i$ is well-defined.

@@

Below, we show that the polynomial $R$ associated to $W$ by Algorithm~\ref{alg WLD to denom via GN} is equal to the denominator $R(W)$ from Definition~\ref{def R(W)}, and is the radical\note{radical} of the product $\prod_{i=1}^n \Delta_{I_i}$. First, however, we give a worked example of Algorithm~\ref{alg WLD to denom via GN}.

@@

\begin{eg}
Consider the Wilson loop diagram in Figure~\ref{fig R eg}. Assigning propagators $p$, $q$, $r$ to rows $1,2,3$ respectively, we obtain the matrix
\[
C(W) = \begin{bmatrix} a & b & 0 & 0 & 0 & c & d \\ e & f & 0 & 0 & g & h & 0 \\ i & j & 0 & k & l & 0 & 0 \end{bmatrix}.
\]
The Grassmann necklace of this diagram is 
\begin{gather*}I_1 = \{1,2,4\},\ I_2 = \{2,4,5\},\ I_3 = \{4,5,6\},\ I_4=\{4,5,6\},\\ I_5=\{5,6,7\},\ I_6 = \{6,7,1\},\ I_7=\{7,1,2\}. \end{gather*}  
Figure~\ref{fig R eg} indicates indicates which vertex is contributed to $I_i$ by each propagator, for each $i \in [1,7]$.

@@

All that remains is to prove that $\Delta_{I_i}$ has no irreducible factors of degree 3 or more.  Suppose for a contradiction that $f$ is a factor of $\Delta_{I_i}$ of degree $\geq 3$. Without loss of generality\footnote{This is without loss of generality for the following reason.  By removing the propagators which come before those contributing to $f$ (in the order given by $I_i$) and changing $i$ to be the first vertex which contributes to $f$, we obtain an admissible diagram.  In this diagram, $f$ still divides $\Delta_{I_i}$ and we also have $i\in I_i$ and that 
$i$ contributes to $f$.  Thus this diagram can be used in place of $W$.} we may assume $i\in I_i$ and that $i$ contributes to $f$.
Finally, we can suppose that $W$ is minimal with respect to number of propagators amongst diagrams with an irreducible factor of degree $\geq 3$.

@@

\textbf{Case 2}: Suppose $p = (i, m)$, i.e. $p$ has an end on the $i^{th}$ edge. If no other propagators are supported on the vertex $i$ then the corresponding column of $C(W)$ has only one nonzero entry in it, and so $\Delta_{I_i}$ has a linear factor contributed by $p$ and $i$; as above, this is a contradiction.
Since $p$ is the first propagator in the ordering imposed by $I_i$, it is the clockwise-most propagator supported on $i$ (by Remark {\ref{clockwise ordering rem}}). Since there are multiple propagators supported on $i$, there must be a $q$ ending on edge $i$, such that $I_i(q) = i+1$, that is second clockwise-most propagators supported by $i$ (see Figure~\ref{fig no big factors}).
Write $q = (i, j)$ for the support of $q$.  The situation for $q$ is very similar to Case 1: in particular, we have $I_{i+1}(q) = j$ by Lemma {\ref{vertex cyclic int lem}} and so if $j\in I_i$ then the propagator which contributes $j$ is outside of $q$.
    %\color{violet}if we have $r$ such that $I_i(r) = j$, as above, $r$ lies outside $q$: $r \in \cP_{out}(q, <_i)$.

@@

\item Suppose propagators $p$ and $q$ are supported on a common edge  $a$, with $I_i(p)=a$ and ${I_i(q)=a+1}$.  Let $x_{p,a},x_{p,a+1},x_{q,a},x_{q,a+1}$ be the associated variables in $C(W)$. For any fixed bijection $\sigma$ from $\cP\setminus\{p,q\}$ to $I_i \setminus\{a,a+1\}$ for which each propagator is supported on its image under the bijection, we can extend $\sigma$ to a bijection from propagators to $I_i$ in two ways: either $p\mapsto a$ and $q\mapsto a+1$ or $p\mapsto a+1$ and $q\mapsto a$.  The sum of the contributions of all these bijections to $\Delta_{I_i}$ is therefore the product of $(x_{p,a}x_{q,a+1}-x_{p,a+1}x_{q,a})$  with the minor coming from $\cP\setminus\{p, q\}$ and $I_i \setminus \{a,a+1\}$.  Since there is no cancellation of terms in the expansion of $\Delta_{I_i}$, if any other terms appear then they will cause a factor which is not in the form described at the beginning of the proof of part (2).  Therefore no such terms exist and $(x_{p,a}x_{q,a+1}-x_{p,a+1}x_{q,a})$ is a factor of $\Delta_{I_i}$.

@@

Taking everything together we have that $R|\prod_{i=1}^n \Delta_{I_i}$ (from part (4)), $R$ contains all factors of $\prod_{i=1}^n \Delta_{I_i}$, and $R$ is square-free.  Therefore the ideal generated by $R$ is the radical of the ideal generated by $\prod_{i=1}^n \Delta_{I_i}$.
  \end{enumerate}
\end{proof}

@@

Thus we have shown a deep relationship between the physically derived poles of the integrand $\Omega(W)$ and the geometry and combinatorics associated to the diagrams $W$. This is yet another piece of substantial evidence indicating that the positroid representation of Wilson loop diagrams is the correct geometric object to use when considering the Wilson loop amplitudes geometrically.
